---
title: "Data Science - Project"
author: "Alexander Prokhorov"
date: "7/14/2021"
output: html_document
---


# Effects of Covid-19 on the populace of Israel

***
# Article summarization and topic background 

The coronavirus disease 2019 (COVID-19) is a contagious disease caused by severe acute respiratory syndrome coronavirus 2 (SARS-CoV-2). The first case was identified in Wuhan, China, in December 2019. It has since spread worldwide, leading to an ongoing pandemic. The first case in Israel was confirmed on 21 February 2020 and on 20 March, an 88-year-old Holocaust survivor in Jerusalem who suffered from previous illnesses was announced as the country’s first casualty. The Israeli Ministry of Health has published daily data on the pandemic, which is open to the general public for reading and reviewing [(https://data.gov.il/dataset/covid-19)] and has been updating data including the number of weekly deaths, the number of tests, the number of patients verified, isolations, the amount of vaccines given in the first and second dose and many more.

Data Comprising Tool: The ANOVA test (Analysis of Variance) is used to compare the mean of multiple groups. ANOVA test is used to investigate differences in means. In this project we will investigate three different research questions using three different types of ANOVA. One-way ANOVA is an extension of the independent sample t-test for comparing means in a situation where there are more than two groups. This is the simplest case of ANOVA test where the data is organized into several groups according to only one single grouping variable. two-way ANOVA used to evaluate simultaneously the effect of two different grouping variables on a continuous outcome variable. three-way ANOVA is used to simultaneously evaluate the effect of three different grouping variables on a continuous outcome variable.

# The Databases

The hypothesis and research questions will attempt to find correlations between
age and gender to mortality, recovery and vaccination from COVID-19 in israel

My databases/libraries are : 
```{r load-data}
library(dplyr)
library(statsr)
library(tidyverse)
library(ggpubr)
library(rstatix)
library(GGally)
library(datarium)
library(emmeans)
library(readr)
library(knitr)

deceased="https://raw.githubusercontent.com/aschlon4/R-Final-Project/main/corona_deceased.csv"
recovered="https://raw.githubusercontent.com/aschlon4/R-Final-Project/main/corona_recovered.csv"
israel="https://raw.githubusercontent.com/aschlon4/R-Final-Project/main/vaccinated-per-day-2021-07-10.csv"

data_vac_per_day_isr <-read_csv(israel)

data_dec <-read_csv(deceased)

data_rec <-read_csv(recovered)

```

## My hypothesis and research questions

In my research I will focus on three different data sets; vaccinated, recovered, and deceased and the effect of age groups and other categorical variables on continuous variables. The first question will be using One Way ANOVA, the second question Two Way ANOVA etc..

Research question 1: Is there a correlation between age groups(categorical variable “age_group”) and the amount of people vaccinated in the first dose (continuous variable “first_dose”)? The data will be taken from the Vaccinated population

Research question 2: Is there a correlation between age groups(categorical variable “Age_group”) and the duration of hospitalization (continuous variable “Length_of_hospitalization”) before death and how does it vary between genders(categorical variable “Gender”)? The data will be taken from the deceased population

Research question 3: Is there a correlation between age groups (categorical variable “age_group”) and the source of infection (categorical variable “test_indication”)to the duration of recovery (continuous variable “days_between_pos_and_recovery”) and how does it vary between genders (categorical variable “gender”)? The data will be taken from the hospitalized population

Question 1:

1) Is there a correlation between age groups(categorical variable “age_group”) and the amount of people vaccinated in the first dose (continuous variable “first_dose”)? The data will be taken from the Vaccinated population
is it different to these genders? 
The relevant database contains a list of deceased patients due to the virus, age group, different times measured such as time from being positive to hospitalization etc..
I used the gender and the time of positive to death to look for a correlation.

Some data manipulation to organize the patients into two groups of gender:
```{r}
# I rearranged the age groups so that they appear in continuous order
levels(data_vac_per_day_isr$age_group)
```

```{r}
data_vac_per_day_isr <- data_vac_per_day_isr %>%
  reorder_levels(age_group, order = c("0-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80-89", "90+"))
```

A summary table of the statistics: 

```{r}
data_vac_per_day_isr %>%
  group_by(age_group) %>%
  get_summary_stats(first_dose, type = "mean")
```
Summary table
```{r}
table1::table1(~ first_dose | age_group , data = data_vac_per_day_isr)
```
Visualization
```{r}
ggboxplot(data_vac_per_day_isr, x = "age_group", y = "first_dose")
```
Judging by the results, we can see that the edges of the age groups, i.e. the youngest and oldest age groups, have the lowest number of vaccinated people while the middle age groups have the highest number of vaccinated people in the first dose.

Outliers
```{r}
data_vac_per_day_isr %>% 
  group_by(age_group) %>%
  identify_outliers(first_dose)
```

There were no extreme outliers.

Normallity assumption
```{r}
namodel  <- lm(first_dose ~ age_group, data = data_vac_per_day_isr)
ggqqplot(residuals(namodel))
```

```{r}
shapiro_test(residuals(namodel))
```
If we check normality assumption by analyzing the model residuals, We can see that in the QQ plot, not all the dots fall along the reference line, and that the edges are off the line, especially at the top-most level.

We might try a different approach and check to see if we can assume normality using the Shapiro-Wilks test for each age group. If the data is normally distributed, the P-value should be greater than 0.05.

```{r}
data_vac_per_day_isr %>%
  group_by(age_group) %>%
  shapiro_test(first_dose)
```
We can see that the data isn’t normally distributed, as the p-value is not greater than 0.05.

```{r}
ggqqplot(data_vac_per_day_isr, "first_dose", facet.by = "age_group")
```
We must remember that if the sample size is greater than 50, the normal QQ plot is preferred because at larger sample sizes the Shapiro-Wilks test becomes very sensitive even to a minor deviation from normality. Our sample size is 1814, which can help explain the lack of correlation between some of the dots to the reference line. Yet, here we can see that most dots do fall into the reference line.

Due to the results, we will perform an AVONA test

Homogeneity of variance assumption
```{r}
plot(namodel, 1)
```
We can see that in the plot above, there are no evident relationships between residual and fitted values (the mean of each "age_group"). Therefore, we can assume the homogeneity of the variances.

We can also use the Levene’s test to check the homogeneity of variances:

```{r}
data_vac_per_day_isr %>% levene_test(first_dose ~ age_group)
```
We can see from the output that the p-value is < 0.05, which means that the null hypothesis is rejected. This means that, there is significant difference between variances across groups. Therefore, we can assume the lack of homogeneity of variances in the different treatment groups.

Computation
```{r}
res.aov <- data_vac_per_day_isr %>% anova_test(first_dose ~ age_group)
```

```{r}
res.aov
```
From the above ANOVA table, we can see that there are significant differences between groups (p = 2.36e-25), which are highlighted with “*“, F(8,1805)=17.519, p=2.36e-25, eta2[g] = 0.072.

Post-hoc test

```{r}
pwc <- data_vac_per_day_isr %>% tukey_hsd(first_dose ~ age_group)
pwc
```
From the output, we can observe that there is a difference between most age groups.

#Report visualization
```{r}
pwc <- pwc %>% add_xy_position(x = "age_group")
ggboxplot(data_vac_per_day_isr, x = "age_group", y = "first_dose") +
  stat_pvalue_manual(pwc, hide.ns = TRUE) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
    )
```
Tukey post-hoc analyses revealed that some age groups were statistically significant from each other while others are not (p <0.0001).We can see that groups "0-19" and "80-89" were similar to each other and differed from the rest, the same applies to age groups "20-29", "30-39" and "70-79", and groups "50-59" and "60-69".Age groups "40-49" and "90+" differ from the rest.

Conclusions:
The various tests and outputs have showed us that the age group of 90+ were the least vaccinated from those who received the first dose, while the largest number of vaccinated people was from the 40-49 age group. The risk group is defined from the age of 60+ so we expected to see more vaccinated people in this age group and above it, but the results did not confirm our expectations. There are several reasons for this, one of which is the average number of people in the different age populations. Although this is not specified in the database, there may be fewer people in the more advanced age groups, which may affect the results. Beyond that, people in a more advanced age group may not have the means to get tested. In addition, the fact that some of the people who vaccinated were people with chronic diseases or conditions that, in turn, suppress the immune system cannot be neglected. Therefore, it is preferable to vaccinate them as well, which can explain the difference in the age of the vaccinated between age groups.

-------------------

Research question 2:

```{r}
str(data_dec)

```

```{r}
data_dec$Gender= as.factor(data_dec$Gender)
data_dec$Length_of_hospitalization = as.integer(data_dec$Length_of_hospitalization)
```
```{r}
question2.data <- data_dec %>% select(Gender, Age_group, Length_of_hospitalization)

# reordering the age groups 
levels(data_dec$Age_group)
```
```{r}
data_dec <- data_dec %>%
  reorder_levels(Age_group, order = c("0-29","30-39","40-49","50-59","60-69", "70-79", "80-89", "90+")) 

# filtering the data
data_dec %>%  select(Gender, Age_group,Length_of_hospitalization) %>%
filter(!is.na(Gender), !is.na(Age_group), !is.na(Length_of_hospitalization)) %>%
  head(n=10)
```
General Note: In some of the code I used the complete data set as is,  for example data_vec_per_day_isr but in some cases like in the outliners test the code did not allow me to display the extreme outliners and I had to group all of the data into variables that are tested in a question, like question2.data for example.
```{r}
data_dec %>%
  group_by(Gender, Age_group) %>%
  get_summary_stats(Length_of_hospitalization, type = "mean")
```
Summary Table
```{r}
table1::table1(~ Length_of_hospitalization |Age_group + Gender, data = data_dec)
```
Summary statistics
```{r}
bxp <- ggboxplot(
  data_dec, x = "Age_group", y = "Length_of_hospitalization",
  color = "Gender", palette = "jco"
  )
bxp
```

we can see that there are no significant differences between the different age groups for the duration of hospitalization and that there is no difference between genders. We can also see that in all age groups most of the outliners are the same, in group 80-89, some of the dots are higher than the rest. 

Assumption Test:

Outliers:
```{r}
question2.data %>%  
  group_by(Gender, Age_group) %>%
  identify_outliers(Length_of_hospitalization) %>%
  head(n=10)
```
There are extreme outliers in each age group.

Normality test for the whole model:
```{r}
model  <- lm(Length_of_hospitalization ~ Gender+Age_group,
             data = data_dec)

ggqqplot(residuals(model))

```
In the QQ plot, not all of the points fall approximately along the reference line, although it seems like most of the points in the middle section do actually match the reference line, the edges does not. We cannot fully assume normality because of that, but we need to remember that the within the age groups there is diversity due to the fact that some patients were healthier than others before hospitalization.

Shapiro Test
```{r}
ggqqplot(data_dec, "Length_of_hospitalization", ggtheme = theme_bw()) +
  facet_grid(Gender ~ Age_group)
```
Most of the points fall approximately along the reference line, for each cell. We might assume normality of the data but we need to take into account that some points in the edges don’t fall along the line.

Homogneity of variance:

```{r}
data_dec %>% levene_test(Length_of_hospitalization ~ Gender*Age_group)
```
The Levene’s test is not significant (p > 0.05). Therefore, we can assume the homogeneity of variances in the different age groups.

ANOVA test

```{r}
res.aov <- data_dec %>% anova_test(Length_of_hospitalization ~ Gender + Age_group)

```
```{r}
res.aov
```
From the above ANOVA table, we can see that there are no significant differences between groups (p = 4.75e-57), which are highlighted with “*“, F(7,6313)=41.145, p=4.75e-57, eta2[g] = 4.4e-02

Post hoc analysis

```{r}
pwc <- data_dec %>% 
  group_by(Gender) %>%
  emmeans_test(Length_of_hospitalization ~ Age_group, p.adjust.method = "bonferroni") 
```
```{r}
pwc
```
Visualizing our results:
```{r}
pwc <- pwc %>% add_xy_position(x = "Gender")
bxp +
  stat_pvalue_manual(pwc) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
    )
```

Conclusions:

In this question we investigated whether there is a relationship between age groups and the duration of hospitalization of those who died of corona and whether there is a gender effect on mortality. we can see that there are no significant differences between the different age groups and the length of hospitalization before death. However, it can be seen that most patients were hospitalized for a period of between 9-19 days. Patients hospitalized for a period longer than 100 days were likely infected with corona while they were already in the hospital and yet aside from those cases, we see that the length of hospitalization is roughly the same in all age groups and all genders, with exceptions in the lower age ranges.

-------------------

Question 3

```{r}
str(data_rec)
```
```{r}
# changing  the variables into factor or integers 
data_rec$gender= as.factor(data_rec$gender)
data_rec$age_group= as.factor(data_rec$age_group)
data_rec$test_indication= as.factor(data_rec$test_indication)
data_rec$days_between_pos_and_recovery = as.integer(data_rec$days_between_pos_and_recovery)

# filtering the data
question3.data <- data_rec %>% select(gender, age_group,test_indication,days_between_pos_and_recovery) %>%
filter(!is.na(gender), !is.na(age_group), !is.na(test_indication), !is.na(days_between_pos_and_recovery)) 
```

```{r}
data_rec %>% sample_n_by(gender, age_group, test_indication, size = 1)
```

```{r}
data_rec %>% 
  group_by(gender, age_group, test_indication) %>%
  get_summary_stats(days_between_pos_and_recovery, type = "mean_sd") 
```
Table

```{r}
table1::table1(~ days_between_pos_and_recovery + test_indication |age_group + gender, data = data_rec)

```

Summary statistics

```{r}
res.aov <- data_rec %>% anova_test(days_between_pos_and_recovery ~ gender + age_group + test_indication)

```
```{r}
res.aov
```
Visualizations

```{r}
levels(data_rec$age_group)

```
```{r}
data_rec <- data_rec %>%
  reorder_levels(age_group, order = c("0-19", "20-29", "30-39", "40-49", "50-59", "60+"))

bxp <- ggboxplot(
  data_rec, x = "age_group", y = "days_between_pos_and_recovery",
  color = "test_indication", palette = "jco", facet.by = "gender"
  )
bxp
```

We can see that there are many outliers in duration of recovery time, this fact can be attributed to the fact that many patients who recovered from corona reported side effects and still continued to experience the symptoms of the disease even though they had already recovered from the virus. In females we can see a much lower rate of infection from abroad in comparison to men.

Outliers

```{r}
question3.data %>%
  group_by(gender,age_group, test_indication) %>%
  identify_outliers(days_between_pos_and_recovery) %>%
head(n=10)
```

There are extreme outliers in each group tested.

Normality Assumption:

```{r}
model  <- lm(days_between_pos_and_recovery ~ gender*age_group*test_indication, data = data_rec)
ggqqplot(residuals(model))
```

```{r}
ggqqplot(data_rec, "days_between_pos_and_recovery", ggtheme = theme_bw()) +
  facet_grid(gender + age_group ~ test_indication)
```

Here we can see that most of the points do correspond to the reference line and both abroad and contact categories are simmilar, while there are still more cases of people who had contact with a confirmed subject. 

Homogneity of variance assumption

```{r}
data_rec %>% levene_test(days_between_pos_and_recovery ~ gender*age_group*test_indication)

```

The Levene’s test is significant (p < 0.05). Therefore, we cannot assume the homogeneity of variances in the different groups.

Computation

```{r}
question3.data  <- data_rec %>%  select(days_between_pos_and_recovery, gender, age_group, test_indication)

res.aov <- question3.data %>% anova_test(days_between_pos_and_recovery ~ gender*age_group*test_indication)
```

```{r}
res.aov
```
There was no statistically significant three-way interaction between gender, test indication and gender to the duration of recovery, p<.05

Post-hoc test

```{r}
model  <- lm(days_between_pos_and_recovery ~ gender*age_group*test_indication, data = data_rec)
data_rec %>%
  group_by(gender) %>%
  anova_test(days_between_pos_and_recovery ~ age_group*test_indication, error = model)
```

Pairwise comparisons

```{r}
pwc <- data_rec %>%
  group_by(gender, age_group) %>%
  emmeans_test(days_between_pos_and_recovery ~ test_indication, p.adjust.method = "bonferroni") %>%
  select(-df, -statistic, -p) 
pwc %>% filter(gender == "M", age_group == "60+")
```
For males at the age 60+ there was a statistically significant mean difference between the test indication “abroad” and “contact with confirmed” to “other” in the duration of recovery.

Visualization: box plots with p-values

```{r}
pwc <- pwc %>% add_xy_position(x = "test_indication")
pwc.filtered <- pwc %>% filter(gender == "M", age_group == "60+")
bxp +
  stat_pvalue_manual(
    pwc.filtered, color = "age_group", linetype = "age_group", hide.ns = TRUE,
    tip.length = 0, step.increase = 0.1, step.group.by = "gender"
  ) +
  labs(
    subtitle = get_test_label(res.aov, detailed = TRUE),
    caption = get_pwc_label(pwc)
    )
```
Conclusions:

In this question we investigated whether there is a relationship between the duration of recovery from corona and the source of infection and whether it differs between age groups and genders. We have seen that there are fewer sources of infection from abroad because the skies were closed for a long period of time, there is a bias in the data. For example, we can see that younger demographics (0-19), who fly less, were less susceptible to contracting the desease by way of flight

There is a slight difference in the recovery time in the group ages of over 20 between men and women, while men infected abroad show a higher recovery time than the women. They also show the longest recovery time among all the groups tested. But aside for that, it can be seen quite well that the duration of recovery is the similar for all age groups and all genders and is not overly affected by the source of infection.
